
---
title: "Exploratory Data Analysis in R"
author: "S. Morteza Najibi, Shiraz University"
date: "April 22, 2019"
#output: beamer_presentation
#theme: "SDATbeamer"
fontsize: 8pt
output: ioslides_presentation
#output: md_document
---


## Graphics in R

  - Exploratory Data Analysis (EDA)  
  - Powerful environment for visualizing scientific data
  - Fully programmable
  - An example of Exploratory Data Analysis will be performed with the Iris data set.  

##  Important high-level plotting functions

  - `plot`: generic x-y plotting
  - `barplot`: bar plots
  - `boxplot`: box-and-whisker plot
  - `hist`: histograms
  - `pie`: pie charts
  - `dotchart`: cleveland dot plots
  - `image`, `heatmap`, `contour`, `persp`: functions to generate image-like plots
  - `qqnorm`, `qqline`, `qqplot`: distribution comparison plots
  - `pairs`, `coplot`, `matplot`: display of multivariant data Help on these functions

## Iris Data

The iris data set is available at the [UCI Machine Learning Repository](http://archive.ics.uci.edu/ml/).

The data set was created by statistician Douglas Fisher.  The data consists of 3 classes of flower types:

  - _setosa_
  - _virginica_
  - _versicolor_

## The Iris Attributes
The data has 4 attributes:

  - sepal width
  - sepal length
  - petal width
  - petal length

## Load Data

The data needs to be loaded into R.

```{r}
head(iris)
names(iris)
```

## Summary Statistics

The summary function can be used to assess the statistical properties of each attribute.

```{r}
summary(iris[,-5])
```
other statistics may be calculated as described in the introduction to R examples.

# Visualizations

## Plot

- plot x versus y
````{r}
plot(iris[,1], iris[,2],xlab=names(iris)[1],ylab=names(iris)[2],col=2)
```

## Some more techniques in `plot`

```{r}
set.seed(100)
y <- matrix(runif(30), ncol=3, dimnames=list(letters[1:10], LETTERS[1:3]))
plot(y[,1], y[,2], pch=20, col="red", main="Symbols and Labels")
text(y[,1]+0.03, y[,2], rownames(y))
```

## Add a regression line

```{r}
plot(y[,1], y[,2])
myline <- lm(y[,2]~y[,1]);
abline(myline, lwd=2)
````

## Add a mathematical expression to a plot
```{r}
plot(y[,1], y[,2])
text(y[1,1], y[1,2]+.2,expression(sum(frac(1,sqrt(x^2*pi)))), cex=1.3)
````

## Histograms

- Set the number of breaks to 9, therefore, there will be 10 equal sized bins.

```{r hist1, out.height='180px'}
par(mfrow=c(1,2))
hist(iris$Petal.Width, breaks=9)
hist(iris$Petal.Width, breaks=19)
```

## some notes
##### Note, the number of breaks can be a single number, vector, function to evaluate points or character string.  

```{r hist3, out.height='150px'}
b <- seq(min(iris$Petal.Width), max(iris$Petal.Width),
            length=11)
hist(iris$Petal.Width, breaks=b, col="blue", xlab="Petal Width",
     main="Histogram of Petal Width")
```

## Density Plots

Density plots can be viewed as smoothed versions of a histogram.  We can estimate the density using R's `density` function

```{r dens1}
dens.pw = density(iris$Petal.Width)
plot(dens.pw, ylab = "f(width)", xlab = "width", main= "Density of Petal Width")
```

## Density plot cont'd
Let's also look at the density function of `petal.length` for each of the three classes of irises.

```{r dens2}
density.setosa = density(iris$Petal.Length[iris$Species == "setosa"])
density.versicolor = density(iris$Petal.Length[iris$Species == "versicolor"])
density.virginica = density(iris$Petal.Length[iris$Species == "virginica"])
```

## Density plot cont'd

```{r dens3}
plot(density.setosa, ylab="f(length)", xlab="length (cm)", main="Density plot of Petal
     Legnths", xlim = c(0,8), lwd=4, col="red")
lines(density.versicolor, col="blue", lwd=4)
lines(density.virginica, col="green", lwd=4)
legend(4, 2.5, c("Setosa", "Versicolor", "Virginica"), lwd=rep(4,3),
       col=c("red", "blue", "green"))
```

## Empirical Cumulative Distribution Functions

The cumulative area under the density function is the cumulative distribution function.  Looking at this function using a sample of points gives use the empirical cumulative distribution function (ECDF).   Let's start by calculating the cumulative areas

```{r cdf1, eval=FALSE}
ecdf.setosa = ecdf(iris$Petal.Length[iris$Species == "setosa"])
ecdf.versicolor = ecdf(iris$Petal.Length[iris$Species == "versicolor"])
ecdf.virginica = ecdf(iris$Petal.Length[iris$Species == "virginica"])
```
The functions may also be plotted (notice the step function patterns)
```{r cdf2, eval=FALSE}
plot(ecdf.setosa, ylab = "F(length)", xlab = "length (cm)", col.h = "red",
     main = "ECDF for Petal Lengths",
      verticals = T, col.v = "red", do.p = F, lwd = 4, xlim = c(0, 8))
lines(ecdf.versicolor, col.h = "blue", col.v = "blue", verticals = T, lwd = 4,
        do.p = F)
lines(ecdf.virginica, col.h = "green", col.v = "green", verticals = T, lwd = 4,
        do.p = F)
```

----

```{r cdf3, eval=TRUE,echo=FALSE, out.height='150px'}
ecdf.setosa = ecdf(iris$Petal.Length[iris$Species == "setosa"])
ecdf.versicolor = ecdf(iris$Petal.Length[iris$Species == "versicolor"])
ecdf.virginica = ecdf(iris$Petal.Length[iris$Species == "virginica"])
plot(ecdf.setosa, ylab = "F(length)", xlab = "length (cm)", col.h = "red",
     main = "ECDF for Petal Lengths",
      verticals = T, col.v = "red", do.p = F, lwd = 4, xlim = c(0, 8))
lines(ecdf.versicolor, col.h = "blue", col.v = "blue", verticals = T, lwd = 4,
        do.p = F)
lines(ecdf.virginica, col.h = "green", col.v = "green", verticals = T, lwd = 4,
        do.p = F)
```

## Quantile Plots

The inverse of the ECDF is the quantile function.  

```{r quantile}
ps = seq(0,1,length=25)
quantile.setosa = quantile(iris$Petal.Length[iris$Species == "setosa"], probs=ps)
quantile.setosa
```

----

```{r quantile2,eval=FALSE}
quantile.versicolor = quantile(iris$Petal.Length
                               [iris$Species == "versicolor"],probs=ps)
quantile.virginica = quantile(iris$Petal.Length
                              [iris$Species == "virginica"], probs=ps)
plot(ps, quantile.setosa, xlab = "p", ylab = "length (cm)",
     main = "Quantiles for Petal Length",
     type="l", lwd = 4, col="red", ylim=c(0,8))
lines(ps, quantile.versicolor, col = "blue", lwd = 4)
lines(ps, quantile.virginica, col = "green", lwd = 4)
```

##


```{r,echo=FALSE}
quantile.versicolor = quantile(iris$Petal.Length[iris$Species == "versicolor"],
                               probs=ps)
quantile.virginica = quantile(iris$Petal.Length[iris$Species == "virginica"],
                              probs=ps)

plot(ps, quantile.setosa, xlab = "p", ylab = "length (cm)", main = "Quantiles for Petal Length",
     type="l", lwd = 4, col="red", ylim=c(0,8))
lines(ps, quantile.versicolor, col = "blue", lwd = 4)
lines(ps, quantile.virginica, col = "green", lwd = 4)
```

## check Normality plot

```{r}
hist(iris$Sepal.Length, probability=TRUE, breaks=15, col="light blue")
rug(jitter(iris$Sepal.Length, 5))
points(density(iris$Sepal.Length), type='l', lwd=3, col='red')
f <- function(t) {
  dnorm(t, mean=mean(iris$Sepal.Length), sd=sd(iris$Sepal.Length) )
}
curve(f, add=T, col="blue", lwd=3, lty=2)
```

## Q-Q plot
A q-q plot is a plot of the quantiles of the first data set against the quantiles of the second data set.
```{r}
x <- iris$Sepal.Length
qqnorm(x)
qqline(x, col="red", lwd=3)
````

## Some more examples

```{r}
par(mfrow=c(1,2))
x <- rnorm(100)
y <- rnorm(100)^2
qqnorm(x, main="Gaussian variable");qqline(x, col="blue", lwd=3)
qqnorm(y, main="Non gaussian variable");qqline(y, col="red", lwd=3)

```

## Box Plots

Box plots are used to compactly show many pieces of information about a variables distribution including some summary statistics.  

A box plot of each attribute of the iris data set

```{r boxplot1}
boxplot(iris[,-1])
```

----

A box plot may also be used for one attribute `petal.length` and show how it varies with the class `Species`

```{r boxplot2}
boxplot(iris$Petal.Length ~ iris$Species)
```

## Scatter Plots

Scatter plots are used to plot two variables against each other (or 3 in the case of 3D plots).

```{r scatter1}
plot(iris$Petal.Length, iris$Sepal.Width)
```

## Pairwise Plots

For data sets with only a few attributes, all the pairwise scatter plots may be constructed.

```{r scatter2}
pairs(as.matrix(iris[,-5]), pch=21, bg=c("red", "blue", "green")[(iris$Species)])
```



## Color Selection Utilities

Default color palette and how to change it

```{r}
palette()
palette(rainbow(5, start=0.1, end=0.2))
palette()
```

----

The gray function allows to select any type of gray shades by providing values from 0 to 1

````{r, out.height='150px'}
pie(1:5,col=gray(seq(0.1, 1, by= 0.2)))
````

## Color gradients with colorpanel function from gplots library

````{r, out.height='100px'}
library(gplots,warn.conflicts = F)
par(mfrow=c(1,2))
pie(1:50, col=colorpanel(50,"red","black","green"))
pie(1:50, col=rainbow(50))
````



## Arranging Plots with Variable Width

The layout function allows to divide the plotting device into variable numbers of rows and columns with the column-widths and the row-heights specified in the respective
arguments.

```{r}
nf <- layout(matrix(c(2,1,3,3), 2, 2, byrow=TRUE), c(3,7), c(5,3))
layout.show(nf)
```

----

```{r,out.height='150px'}
x <- rnorm(100) ; y <- rnorm(100)
layout(matrix(c(2,1,3,3), 2, 2, byrow=TRUE), c(3,7), c(5,4))
plot(x,y)
boxplot(y,horizontal = FALSE,col=2)
hist(x,col=4)
````

## Margine plots

````{r,eval=FALSE}
data(cars)
layout( matrix( c(2,1,0,3), 2, 2, byrow=T ),
c(1,6), c(4,1))
par(mar=c(1,1,5,2)); plot(cars$dist ~ cars$speed,xlab='', ylab='',las = 1)
rug(side=1, jitter(cars$speed, 5) );rug(side=2, jitter(cars$dist, 20) );
title(main = "cars data")
par(mar=c(1,2,5,1))
boxplot(cars$dist, axes=F); title(ylab='Stopping distance (ft)', line=0)
par(mar=c(5,1,1,2))
boxplot(cars$speed, horizontal=T, axes=F); title(xlab='Speed (mph)', line=1)
```

----

````{r,eval=TRUE, echo=FALSE, out.height='150px'}
data(cars)
layout( matrix( c(2,1,0,3), 2, 2, byrow=T ),
c(1,6), c(4,2))
par(mar=c(1,1,5,2)); plot(cars$dist ~ cars$speed,xlab='', ylab='',las = 1)
rug(side=1, jitter(cars$speed, 5) );rug(side=2, jitter(cars$dist, 20) );
title(main = "cars data")
par(mar=c(1,2,5,1))
boxplot(cars$dist, axes=F); title(ylab='Stopping distance (ft)', line=0)
par(mar=c(5,1,1,2))
boxplot(cars$speed, horizontal=T, axes=F); title(xlab='Speed (mph)', line=1)
```


## Exercise: Bar Plots, Pie plot,

  - Consider the same approach in this slides for `barplot`

## Saving Graphics to Files

After the pdf() command all graphs are redirected to file test.pdf. Works for all common formats similarly:
jpeg, png, ps, tiff, ...

```{r,eval=FALSE}
pdf("test.pdf"); plot(1:10, 1:10); dev.off()
```


# 3 Dimesional functions and their plots

## `persp` 3d plot

```{r,out.height='100px'}
x <- seq(-10, 10, length= 30)
y <- x
f <- function(x, y) { r <- sqrt(x^2+y^2); 10 * sin(r)/r }
z <- outer(x, y, f)
persp(x, y, z, theta = 30, phi = 30, expand = 0.5, col = "lightblue")
```

## Countour Plots

A contour plot is a graphical technique for representing a 3-dimensional surface by plotting constant z slices, called contours, on a 2-dimensional format.
```{r,out.height='100px'}
contour(x,y,z) # or contour(z)
```

## 2D Density

Density estimation is available for higher dimensional data.  Use the `MASS` package.

```{r contour}
library(MASS)
petal.dens = kde2d(iris$Petal.Length, iris$Petal.Width)
contour(petal.dens)
```

----

The plot may aslo be viewed as an image

```{r contour2}
image(petal.dens)
```


# lattice Grapphics

## What is lattice?
- High-level graphics system
- Implements Trellis graphics system from S-Plus Simplifies high-level plotting tasks: arranging complex graphical features
- Syntax similar to R’s base graphics

## Scatter Plot Sample

```{r,out.height='100px'}
library(lattice)
xyplot(1:8 ~ 1:8 | rep(LETTERS[1:4], each=2), as.table=TRUE)
```

## Line Plot Sample

```{r}
parallelplot(~iris[1:4] | Species, iris, horizontal.axis = FALSE,
             layout = c(1, 3, 1))
```


# ggplot2 package (a superfull package in graphic )
## ggplot2

  * What is ggplot2?
      + High-level graphics system
      + Streamlines many graphics workflows for complex plots
      + Simpler qplot function provides many shortcuts

  * ggplot function accepts two arguments
      + Data set to be plotted
      + `Aesthetic` mappings provided by `aes` function
      + Additional parameters such as geometric objects (e.g. points, lines, bars) are passed on by appending them with + as separator.

## qplot function

qplot syntax is similar to R’s basic plot function Arguments:

- x: x-coordinates (e.g. col1)
- y: y-coordinates (e.g. col2)
- data: data frame with corresponding column names xlim, ylim: e.g. xlim=c(0,10)
- log: e.g. log="x" or log="xy"
- main: main title; see ?plotmath for mathematical formula xlab, ylab: labels for the x- and y-axes
- color, shape, size
- ...: many arguments accepted by plot function

----

Create sample data
```{r}
library(ggplot2,quietly = T)
#cat <- rep(c("A", "B"), 5)
qplot(iris$Sepal.Length, iris$Sepal.Width, geom="point")
````

## Prints dots with different sizes and colors

```{r}
qplot(Sepal.Length, Petal.Width, data = iris, geom="point",
      size=Petal.Length,col=Species,
      main="Dot Size and Color Relative to Some Values")
```


## Drops legend

```{r}
qplot(Sepal.Length, Petal.Width, data = iris, geom="point",col=Species)+
       theme(legend.position = "none")
```



## Plot different shapes

```{r}
qplot(Sepal.Length, Petal.Width, data = iris, geom="point",
      shape=iris$Species,size=2)
```

## Scatter Plot with Regression Line

```{r}
qplot(Sepal.Length, Petal.Width, data = iris, geom = c("point"))+
            geom_smooth(method=lm,col=2)
```

## More application of qplot

```{r}
qplot(Sepal.Length, Petal.Width, data = iris, facets = Species ~ Petal.Length)
```
